Making Sequential Programs Deterministic

### 2019-04-06 design.tex
Finish draft of section 2.4.3

This section introduces the another polymorphism feature by contrasting the
types of the \<get> and \<set> operations on lists.

@@

$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$

\noindent
Its instantiations are:

@@

@@

\noindent
It cannot be instantiated at $\kappa = \<OrderNonDet>$, but the other two
instantiations are enough to ensure that if either the list or the index is (order) non-deterministic, then the result is
nondeterministic.

@@

A similar typing does \emph{not} work for the \<set> function that sets an
element of a list, because it would permit a deterministic collection to
be side-effected in non-deterministic ways.  Consider the following
possible type for

$\<set> : \forall \forallt{\tau}$

qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

>>>>

This section introduces another polymorphism feature by contrasting the
types of the \<get> and \<set> operations on lists.

@@

$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \up{\kappa}\ \<Int> \rightarrow \up{\kappa}\ \<String>$

\noindent
(The actual type permits the qualifier on the type argument to differ
from the type qualifier on the list, and it better handles
order-nondeterminism, but this type is enough for illustration in this section.)

\todo{Show actual type.}

Its instantiations are:

@@

$\<get> : \<OrderNonDet List>\angles{\<Det String>} \times \<NonDet Int> \rightarrow \<NonDet String>$

@@

\noindent
The type of \<get> ensures that if either the list or the index is (order)
non-deterministic, then the result of the call is nondeterministic.

@@

A similar typing does \emph{not} work for the \<set> function that sets an
element of a list.  Consider the putative type

$\<set> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \up{\kappa}\ \<Int> \times \down{\kappa}\ \<String> \rightarrow ()$

\noindent
which has instantiations

$\<set> : \<NonDet>\ \<List>\angles{\<NonDet>\ \<String>} \times \<NonDet>\ \<Int> \times \<NonDet>\ \<String> \rightarrow ()$

$\<set> : \<OrderNonDet>\ \<List>\angles{\<Det>\ \<String>} \times \<NonDet>\ \<Int> \times \<Det>\ \<String> \rightarrow ()$

$\<set> : \<Det>\ \<List>\angles{\<Det>\ \<String>} \times \<Det>\ \<Int> \times \<Det>\ \<String> \rightarrow ()$

\noindent
Suppose a client calls this method as follows:
\begin{Verbatim}[commandchars=\\\{\}]
    Det \lptype{List}{Det String} dList = ...
    NonDet int random = ...
    Det String str = ...
    set(dList, random, str)  // should be forbidden
\end{Verbatim}

\noindent
The call should be forbidden, since on different runs it may make changes at different
indices of a deterministic list.
However, the call is permitted.  The middle instantiation is applicable, since
$\<Det> \ptype{List}{Det String} \sqsubseteq \<OrderNonDet>
\ptype{List}{Det String}$ and $\<Det> \sqsubseteq \<NonDet>$.

\OurTypeSystem must prevent this behavior.  It does so via another variant
of qualifier variables, $\use{\kappa}$, which represents a \emph{use} of
$\kappa$ that does not affect the \emph{instantiation} of $\kappa$.

Ordinarily, the polymorphic function is instantiated at the least upper
bound of the types of all the arguments that correspond to uses of the type
parameter.
For example, function

$f : \forallt{\kappa} \kappa\ \<Int> \times \<Det>\ \<Int> \times\kappa\ \<Int> \times \kappa\ \<Int>$

\noindent
would be instantiated at the least upper bound of the types of its first,
third, and fourth arguments at a given call.  (If no such instantiation
exists, or if the other arguments do not conform to the parameter types, the call does not type-check.)
By contrast, function

$f : \forallt{\kappa} \kappa\ \<Int> \times \<Det>\ \<Int> \times\use{\kappa}\ \<Int> \times \kappa\ \<Int>$

\noindent
would be instantiated at the least upper bound of the types of its first
and forth arguments, and the third argument must conform to the
instantiation.  However, the third argument does not affect the instantiation.

This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver.
The correct type for \<set> on \ptype{List}{String} is

$\<set> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \use{\kappa}\ \<Int> \times \down{\kappa}\ \<String> \rightarrow ()$

\noindent
and the undesired call \<set(dList, random, str)> does not type-check.

(For the curious reader, the actual types of the \<get> and \<set> methods
in the JDK are the following.

\begin{Verbatim}
    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
    @PolyDet("up") E set(@PolyDet List<E> this, @PolyDet("use") int index, E element);
\end{Verbatim}

\noindent
The syntax is explained in \cref{java-polymorphism}.)

### 2019-04-06 randoop-case-study.tex
Finish draft of section 2.4.3

The number of \<@Det> annotations --- one per 22 lines of code --- is much
higher than we would prefer.
Nonetheless, it compares favorably to the
extensive effort by the Randoop developers (\cref{sec:subject-program}),
especially given that that \theDeterminismChecker found issues that the
developers had not.

>>>>

The number of \<@Det> annotations --- one per 22 lines of code --- is much
higher than we would prefer.
Nonetheless, it compares favorably to the
extensive effort by the Randoop developers (\cref{sec:subject-program}),
especially given that that \theDeterminismChecker found issues that the
developers did not.
As another point of comparison,
there are fewer type qualifiers than Java generic type arguments.
